---
title: "Class 7: Machine Learning 1"
author: "Nathan Joseph (PID: A17668656)"
format: pdf
toc: true
---

## Background
Today we will begin our exploration of some important machine learning methods, namely **clustering** and **dimensionality reduction**.

Let's make up some input data for clustering for we know what the natural "clusters" are.

The function `rnorm()` can be useful here..

```{r}
hist(rnorm(50, mean=10))
```

> Q. Generate 30 random numbers centered at +3 and another 30 centered at -3

```{r}
tmp <- c(rnorm(30, mean=3), rnorm(30, mean=-3))

x<-cbind(x=tmp, y=rev(tmp)) # columwise binding

plot(x)
```



## K-means clustering

The main function in "base R" for K-means clustering is called `kmeans()`:

```{r}
km <- kmeans(x, centers=2)
```


> Q. What component of the results object details the cluster sizes?

```{r}
km$size
```
> Q. What component of the results object details the cluster centers?

```{r}
km$centers
```

> Q. What component of the results object details the cluster membership vectors (i.e. our main result of which points lie in which cluster)?

```{r}
km$cluster
```

> Q. Plot our clustering results with points colored by cluster and also add the cluster centers as new points colored in blue?

```{r}
plot(x, col=c(km$cluster))
points(km$centers, col='blue', pch=20)
```

> Q. Run `kmeans()` again and this time produce 4 clusters (call your result object `k4`) and make a results figure like above?

```{r}
k4 <- kmeans(x, 4)
plot(x, col=k4$cluster)
points(k4$centers, col='purple', pch=20)
```


The metric
```{r}
km$tot.withinss
k4$tot.withinss
```

> Q. Let's try different number of K (centers) from 1 to 30 and see what the best result is?

```{r}
ans <- NULL
for (i in 1:30) {
  ans <- c(ans, kmeans(x, centers = i)$tot.withinss)
}
ans
```

```{r}
plot(ans, typ="o")
```

Note: Although it is possible to cluster up to 60 centers, it isn't optimal since the elbow or knee point is at 2 which is the optimal k for clustering.

**Key-point:** K-means will impose a clustering structure on your data even if it is not there - it will always give you the answer you asked for even if that answer is silly!


## Hierarchical Clustering

The main function of Hierarchical Clustering is called `hclust()`. Unlike `kmeans()` (which does all the work for you) you can't just pass `hclust()` our raw input data. It needs a "distance matrix" like the one returned from the `dist()` function.


```{r}
d<-dist(x)
hc<-hclust(d)
plot(hc)
```

To extract our cluster membership vector from a `hclust` result object we have to "cut" our tree at a given height to yield separate "groups"/"branches".
```{r}
plot(hc)
abline(h=8, col='red', lty=2)
```

To do this we use the `cutree()` function on our `hclust()` object:

```{r}
groups <- cutree(hc, h=8)
groups
```

```{r}
table(groups, km$cluster)
```

## PCA of UK food data

Import the dataset of food consumption in the UK:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
nrow(x)
```

One solution to set the row names is to do it by hand...

```{r}
rownames(x) <- x[,1]
x
```

To remove the first column, I can use the minus index trick 

```{r}
x <- x[,-1]
x
```

A better way to do this is to set the row names to the first column with `read.csv()`

```{r}
x <-read.csv(url, row.names=1)
x
```


>Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer the second way since it is more robust and it will not interfere with my future analysis with this dataset.

### Spotting major differences and trends

Is difficult even in this wee 17D dataset...

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))

```

### Pairs plots and heatmaps

> Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

If a given point is on the diagonal then the two categories are the same for the 2 countries in the plot.

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

```{r}
library(pheatmap)
```


```{r}
pheatmap( as.matrix(x) )
```

## PCA to the rescue

The main PCA function in "base R" is called `prcomp()`. This function wants the transpose of our food data as input (i.e. the food as columns and the countries as rows).

```{r}
pca <- prcomp(t(x))
```

```{r}
summary(pca)
```
```{r}
attributes(pca)
```

To make one of main PCA result figures we turn to `pca$x` the scores along our new PCs. This is called "PC plot" or "score plot" or "Ordination plot" ...

```{r}
my_cols <- c("orange", "red", "blue", "darkgreen")
```


```{r}
library(ggplot2)
ggplot(pca$x) +
  aes(PC1, PC2, col=my_cols) +
  geom_point()
```

The second major result figure is called a "loadings plot" of "variable contributions plot" or "weight plot".

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```


